\documentclass{article}

\usepackage{amsmath}
\usepackage{siunitx}
\usepackage{tikz}
\usepackage{booktabs}
\usepackage{bm}
\usepackage{xcolor}
\usepackage[margin=0.75in]{geometry}
\usepackage[calc, showdow,english]{datetime2}
\usepackage[backend=biber, style=ieee]{biblatex}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\addbibresource{references.bib}

\algnewcommand{\Initialise}[1]{
  \State \textbf{Initialise:}
  \Statex \hspace*{\algorithmicindent}\parbox[t]{.8\linewidth}{\raggedright #1}
}

\DTMnewdatestyle{mydateformat}{
  \renewcommand{\DTMdisplaydate}[4]{
    \DTMshortweekdayname{##4},\space
    \DTMmonthname{##2}\nobreakspace
    \number##3,\space
    \number##1
  }
  \renewcommand{\DTMDisplaydate}{\DTMdisplaydate}%
}

\title{Underwater Acoustic Simulation Theory}
\author{mv-2}
\date{\today}

\begin{document}
\maketitle
\tableofcontents

\section{Background \& Motivation}
This project is intended as a learning exercise for both the physics of sound propagation and the programming aspects required for this project. \\
The intended theory to be explored is initially to develop ray propagation simulation based on repeated application of Snell's law and then to layer on Gaussian Beam Tracing (GBT) to eventually develop a well rounded sound prediction tool for higher frequency acoustics. If this is successful, a nodal based solver for low frequency noise will be developed although this would be a large undertaking and is a stretch goal at this time.

\section{Geometric Ray Tracing}
\subsection{Physical Theory}
The geometric ray trace theory used in this simulation is based on work from \textcite{Hovem13}. This methodology discretises the sound speed profile into evenly spaced layers and applies Snell's law (Equation \ref{eq:snell}) across each layer in order to determine the ray path. 
\begin{equation}
  \label{eq:snell}
  \frac{\sin{(\theta_1)}}{\sin{(\theta_2)}}=\frac{c_1}{c_2}
\end{equation}
By assuming that the ray is only refracted between all fluid layers across the entire solution space then this relationship can be reduced to the parameter $\xi$ as shown in Equation \ref{eq:ray_param}.
\begin{align}
  \label{eq:ray_param}
  \xi = \frac{\cos{(\theta(z))}}{c(z)} = \frac{\cos{(\theta_0)}}{c_0}
\end{align}
For a full derivation see \textcite{Hovem13}. For the purposes of the acoustic simulation, Equations \ref{eq:dr_int} and \ref{eq:dt_int} are required.
\begin{equation}
  \label{eq:dr_int}
  \Delta r = \int^{z_2}_{z_1}\frac{\xi c(z)}{\sqrt{1-\xi^2\cdot c^2(z)}}dz
\end{equation}
\begin{equation}
  \label{eq:dt_int}
  \Delta t = \int^{z_2}_{z_1}\frac{1}{c(z)\sqrt{1-\xi^2\cdot c^2(z)}}dz
\end{equation}
To compute these intervals the function for $c(z)$ must be known. Based on the form of the integrals in Equations \ref{eq:dr_int} and \ref{eq:dt_int} not all function forms allow for the integrals to be solved analytically. The most obvious and consequently useful form of $c(z)$ is to assume it is a set of piecewise linear functions along each layer of the fluid medium. Assuming that a general function is known for $c(z)$ either analytically or from curve fitting then the fluid layer $i$ should have the form shown in Equation \ref{eq:layer_lin}.
\begin{equation}
  \label{eq:layer_lin}
  \begin{split}
    c_i(z) = c_i+g_i\cdot(z-z_i),\;\mathrm{for}\;z\in[z_i, z_{i+1}) \\
    \mathrm{where}\;g_i=\frac{c_{i+1}-c_{i}}{\Delta z}
  \end{split}  
\end{equation}
Solving the integrals with this form of sound speed profile leads to the relationships shown in Equations \ref{eq:dr} and \ref{eq:dt}.
\begin{equation}
  \label{eq:dr}
  \Delta r = \frac{\sqrt{1-\xi^2\cdot c^2(z_i)} - \sqrt{1-\xi^2\cdot c^2(z_{i+1})}}{\xi\cdot g_i}
\end{equation}

\begin{equation}
  \label{eq:dt}
  \Delta t = \frac{\ln{\left( \frac{c(z_{i+1})}{c(z_i)}\frac{1 + \sqrt{1 - \xi^2\cdot c^2(z_i)}}{1 + \sqrt{1 - \xi^2\cdot c^2(z_{i+1})}} \right)}}{|g_i|}
\end{equation}
Note that upon inspection of these equations it is clear that if the term $\xi\cdot c(z_{j})\geq 1$ then the equations will both return imaginary quantities. This is non-physical behaviour and must be handled as an edge case. If the increment in range or time is imaginary then it can be concluded that the ray does not reach the next layer of the fluid medium and instead reverses depth direction and $z_{i+1}=z_i$. This lays the groundwork for the ray tracing algorithms although some extra details need to be considered such as how the sound speed profile is to be expressed and other programmatic elements.

\subsection{Ray Tracing Algorithm}

\begin{algorithm}
  \caption{Ray tracing algorithm}\label{alg:ray}
  \begin{algorithmic}
    \Initialise{\strut$r_0\gets \mathrm{source range}$ \\$z_0\gets \mathrm{source depth}$ \\$\theta_0\gets \mathrm{source ray angle}$ \\$z_{\mathrm{sign}}=\mathrm{sign}(\sin(\theta_0))$ \\$i\gets 0$ \\$t\gets 0$\\$maximum iterations$}
    \State $c_0\gets c(z_0)$, $c_1\gets c(z_0+z_{\mathrm{sign}}\cdot\Delta z)$
    \State $\xi\gets \frac{\cos(\theta_0)}{c_0}$
    \While{$i<maximum iterations$}{
      $g_i=\frac{c_{i+1} - c_i}{\Delta z}$
      \If {$\xi\cdot g_i < 1$}
        \State $\Delta r = \frac{\sqrt{1-\xi^2\cdot c^2(z_i)} - \sqrt{1-\xi^2\cdot c^2(z_{i+1})}}{\xi\cdot g_i}$ 
        \State $\Delta t = \frac{\ln{\left( \frac{c(z_{i+1})}{c(z_i)}\frac{1 + \sqrt{1 - \xi^2\cdot c^2(z_i)}}{1 + \sqrt{1 - \xi^2\cdot c^2(z_{i+1})}} \right)}}{|g_i|}$
        \State $z_{i+1}\gets z_i+z_{\mathrm{sign}}\cdot\Delta z$
      \Else
        \State $\Delta r=\frac{2\cdot\sqrt{1-\xi^2\cdot c^2(z_i)}}{\xi\cdot g_i}$
        \State $\Delta t=\frac{2\cdot\ln{(\frac{1 + \sqrt{1-\xi^2\cdot c^2(z_i)}}{\xi\cdot c(z_i)})}}{|g_i|}$
        \State $z_{\mathrm{sign}}\gets-z_{\mathrm{sign}}$
      \EndIf
      \State $r_{i+1}\gets r_i+\Delta r$
      \State $t_{i+1}\gets t_i+\Delta t$
      \State $i\gets i+1$
    }
  \end{algorithmic}
\end{algorithm}

\nocite{*}
\printbibliography

\end{document}
